---
title: "ogttMetrics -- methods for OMNICarb OGTT data analysis"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{ogttXover}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::pdf_document:
    toc: yes
    number_sections: yes
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

# Introduction

This report and software is motivated by experience
with the use 
of oral glucose tolerance tests in an
experimental study of nutritional strategies
to reduce the risk of diabetes and cardiovascular disease (cite
OMNICarb).  

Briefly, a factorial cross-over design was employed
to assess the effects of varying carbohydrate volume (C vs. c
denoting high vs. low carbohydrate content) and
glycemic index (G vs. g denoting high vs. low glycemic index).
Each person in the study consumed a random permutation
of the sequence (CG, Cg, cG, cg), with, for example, "CG" denoting a diet with
high carbohydrate content, high glycemic index.


# Setup

A small segment of code (hidden) creates a container
for all the OGTT studies collected in the study.
```{r setup, echo=FALSE}
suppressPackageStartupMessages({
library(MultiAssayExperiment)
library(ogttMetrics)
library(Biobase)
library(SummarizedExperiment)
library(dplyr)
})
data(xoverSamp)
xoverSamp
```

Missing data in the crossover study was limited.

```{r defineupst,echo=FALSE,results="hide"}
upsetSamples <- function(MultiAssayExperiment,
                         nsets=length(MultiAssayExperiment),
                         nintersects = 24, order.by = "freq",
                         idclip = function(x) substr(x, 1, 12), ... ) { 
    maesn <- colnames(MultiAssayExperiment)
    st <- idclip(maesn[[1L]])
    for (i in seq_along(maesn)[-1])
        st <- union(st, idclip(maesn[[i]]))
    nr <- length(st)
    incid <- matrix(0, nrow = nr, ncol = length(maesn))
    rownames(incid) <- as.character(st)
    for (i in seq_along(maesn))
        incid[, i] <- 1*(rownames(incid) %in% idclip(maesn[[i]]))
    colnames(incid) <- names(MultiAssayExperiment)
    UpSetR::upset(data.frame(incid), nsets = nsets, nintersects = nintersects,
                  sets = colnames(incid), order.by = order.by, ...)
}
```

```{r lkup,fig=TRUE, height=3.5, width=6}
upsetSamples(subsetByAssay(xoverSamp, 
  paste0("insulin_", c("base", "CG", "Cg", "cG", "cg"))), mb.ratio=c(.5,.5))
```
 
# Adding diet-specific metrics

## Matsuda's index

A widely used measure of insulin sensitivity derivable from
OGTT is the Matsuda index.  We use the formula provided
at http://mmatsuda.diabetes-smc.jp/english.html, in the "Web calculator",
for the short OGTT protocol.

```{r domats1}
data(obaSamp)
omnilit = subsetByAssay(obaSamp, c("insulin", "glucose"))
omnilit = addMatsuda120(omnilit)
omnilit
```

## Baseline weight

```{r dowgt}
w = matrix(pData(omnilit)$bodyweight,nr=1)
rownames(w) = "bodyweight"
colnames(w) = rownames(pData(omnilit))
newEL = ExperimentList(c(experiments(omnilit)@listData, list(bodyweight=w)))
omnilit = MultiAssayExperiment(newEL, pData(omnilit)) # set up sampleMap
metadata(omnilit) = list(times=c(0, 10, 20, 30, 60, 90, 120))
omnilit = as(omnilit, "ogttCohort")
```

## Insulin sensitivity

### For a single OGTT series

```{r lkmm}
mm = minmodByID(omnilit, "1")
names(mm)
mm$fit
```

### Visualization of the fit

```{r domo,fig=TRUE}
plot_OGTT_fit(mm)
```

### For a cohort

Here we demonstrate fitting of the minimal model
to estimate $S_I$ (insulin sensitivity).  We also
illustrate how to update a multi-assay experiment
object and coerce to ogttCohort instance, which includes
formal representation of assay times.

```{r docoh, cache=TRUE}
library(parallel)
options(mc.cores=3)
allid = rownames(pData(omnilit)) 
ans = mclapply(allid, function(x) { try(minmodByID( omnilit, x))})
SIs = sapply(ans, function(x) 
   { 
   if(inherits(x, "try-error")) return(NA) else coef(x$fit)["SI"]
   })
summary(SIs)
SIs = matrix(SIs, nr=1)
rownames(SIs) = "SI"
colnames(SIs) = rownames(pData(omnilit))
newEL = ExperimentList(c(experiments(omnilit)@listData, list(SI=SIs)))
omnilit = MultiAssayExperiment(newEL, pData(omnilit)) # set up sampleMap
metadata(omnilit) = list(times=c(0, 10, 20, 30, 60, 90, 120))
omnilit = as(omnilit, "ogttCohort")
```

This process has been carried out on the data from baseline and four
diets, using the following (unevaluated) code pattern, in which the initial
versions of data possess only insulin, glucose and bodyweight
assay data.

```{r doconv,eval=FALSE}
omnicBase = addMatsuda120(updateOGTT(omnicBase))
library(parallel)
options(mc.cores=3)
siBase = matrix(getMinmodSIs(omnicBase, iter=mclapply), nr=1)
rownames(siBase) = "SI"
omnicBase = addAssay(omnicBase, siBase, "SI")
```

# Proceeding to multivariate analysis

Here we work with a simple subset of the full OMNICarb data.

We can concatenate data from multiple treatments
using the concat function.

```{r doconc,eval=TRUE}
data(omnicBase_samp)
omnicBase_samp = addMatsuda120(omnicBase_samp, gname="glucose_base",
        iname="insulin_base")
data(omnicbCbG_samp)
omnicCG_samp = addMatsuda120(omnicCG_samp, gname="glucose_CG",
        iname="insulin_CG")
data(omniclclg_samp)
omniccg_samp = addMatsuda120(omniccg_samp, gname="glucose_cg",
        iname="insulin_cg")
data(omnicbClg_samp)
omnicCg_samp = addMatsuda120(omnicCg_samp, gname="glucose_Cg",
        iname="insulin_Cg")
data(omniclcbG_samp)
omniccG_samp = addMatsuda120(omniccG_samp, gname="glucose_cG",
        iname="insulin_cG")
allmsi = concat(list(base=omnicBase_samp, CG=omnicCG_samp,
    Cg=omnicCg_samp, cG=omniccG_samp, cg=omniccg_samp))
allmsi
```

We will obtain a 'long thin' representation, decode
the diet, and obtain a paired t-test for a contrast of interest.
```{r getp1,eval=TRUE}
msiThin = rearrange(allmsi)
diet = gsub(".*_", "", msiThin$assay)
msiThin$diet = diet
library(dplyr)
library(magrittr)
CG = data.frame(msiThin) %>% 
   filter(diet=="CG" & rowname=="Mats120") %>% 
   rename(mat_CG=value)
cg = data.frame(msiThin) %>% 
     filter(diet=="cg" & rowname=="Mats120") %>% 
     rename(mat_cg=value)
t1 = inner_join(CG, cg, by="primary")
with(t1, t.test(mat_cg-mat_CG, var.equal=TRUE, conf.level=.99))
```

We can work flexibly from the long thin representation to
reproduce JAMA figure 3 element on insulin sensitivity.
```{r docodeth,eval=TRUE}
merge_diets = function(lth, d1="CG", d2="cg", measure="Mats120") {
  di1 = data.frame(lth) %>%
         filter(diet==d1 & rowname==measure) %>%
         rename(ocd1 = value)
  di2 = data.frame(lth) %>%
         filter(diet==d2 & rowname==measure) %>%
         rename(ocd2 = value)
  inner_join(di1, di2, by="primary")
}
cgCG = merge_diets(msiThin, d1="cg", d2="CG")
with(cgCG, t.test(ocd1-ocd2, var.equal=TRUE, conf.level=.99))
#CgCG = merge_diets(msiThin, d1="Cg", d2="CG")
#with(cgCG, t.test(ocd1-ocd2, var.equal=TRUE, conf.level=.95))
generateTests = function(lth, pairs, levels, measure="Mats120") {
  allm = lapply(pairs, function(x) merge_diets(lth, d1=x[1], d2=x[2], measure=measure))
  ans = lapply(seq_along(allm), function(x) with(allm[[x]], t.test(ocd1-ocd2,
   var.equal=TRUE, conf.level=levels[x])))
  nms = sapply(pairs, paste, collapse="-")
  names(ans) = nms
  ans
}
fig3tests = generateTests(msiThin, 
   list(c("Cg", "CG"), c("cg", "cG"), 
        c("cG", "CG"), c("cg", "Cg"), c("cg", "CG")), 
   levels=c(.95, .95, .95, .95, .99))
```

Now we can generate an analog of Figure 3, based however
on the subset in the package.  Contact the maintainer
for the complete dataset.

```{r dof3, fig=TRUE, echo=FALSE, eval=TRUE}
ints = sapply(fig3tests, "[[", "conf.int")
mns = sapply(fig3tests, "[[", "estimate")
nint = length(mns)
xl = range(as.numeric(ints))*1.05
par(mar=c(4,5,2,2))
plot(mns, nint:1, pch=19, axes=FALSE, 
    xlab=expression(paste(Delta, " Matsuda")), xlim=xl,
    ylim=c(.8,nint+.2), ylab=" ")
segments(ints[1,], nint:1, ints[2,], nint:1)
axis(1)
axis(2, at=nint:1, labels=names(fig3tests),las=2)
abline(v=0)
```

```{r dof3b, fig=FALSE, echo=FALSE, eval=FALSE}
fig3btests = generateTests(msiThin, 
   list(c("Cg", "CG"), c("cg", "cG"), 
        c("cG", "CG"), c("cg", "Cg"), c("cg", "CG")), 
   levels=c(.95, .95, .95, .95, .99), measure="SI")
intsb = sapply(fig3btests, "[[", "conf.int")
mnsb = sapply(fig3btests, "[[", "estimate")
nintb = length(mnsb)
xlb = range(as.numeric(intsb))*1.05
par(mar=c(4,5,2,2))
plot(mnsb, nintb:1, pch=19, axes=FALSE, 
    xlab=expression(paste(Delta, " SI")), xlim=xlb,
    ylim=c(.8,nintb+.2), ylab=" ")
segments(intsb[1,], nintb:1, intsb[2,], nintb:1)
axis(1)
axis(2, at=nintb:1, labels=names(fig3btests),las=2)
abline(v=0)
```

Upon adding minimal model fits, the following code
could be used to produce analogous interval displays for SI.

```{r finalpl,fig=FALSE,eval=FALSE}
par(mfrow=c(2,2), mar=c(4,5,2,2))
plot(mns, nint:1, pch=19, axes=FALSE, 
    xlab=expression(paste(Delta, " Matsuda")), xlim=xl,
    ylim=c(.8,nint+.2), ylab=" ")
segments(ints[1,], nint:1, ints[2,], nint:1)
axis(1)
axis(2, at=nint:1, labels=names(fig3tests),las=2)
abline(v=0)
plot(mnsb, nintb:1, pch=19, axes=FALSE, 
    xlab=expression(paste(Delta, " SI")), xlim=xlb,
    ylim=c(.8,nintb+.2), ylab=" ")
segments(intsb[1,], nintb:1, intsb[2,], nintb:1)
axis(1)
axis(2, at=nintb:1, labels=names(fig3btests),las=2)
abline(v=0)
```
